---
title: "MB5370 R Project"
author: "Katie Dyck"
date: "2026-08-13"
format:
  html:
      df-scroll: true
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      code-overflow: scroll
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Housekeeping

Purge the global environment with the command `rm(list = ls())`.

```{r Clear session environment, echo=FALSE, include=FALSE}
#| output: false

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```

# Load packages

```{r load packages}
library(usethis)
library(tidyverse)
library(hms)
library(sf)
library(rnaturalearth)
library(plotly)
```

# Data

## Tracking data

```{r import tracking data}
whales <- read_csv("C:/Users/jc984308/Downloads/AAS_4600_EastAus_humpback_satellite_tracking_data/AAS_4600_EastAus_humpback_satellite_tracking_data/1.Argos_locs.csv")
#whales <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/1.Argos_locs.csv")
glimpse(whales)
```

```{r rename columns}
# rename columns
whales <- whales |> 
    rename(PTT = `Argos PTT`,
           argos_loc_class = `Argos location class`,
           lon = Longitude, lat = Latitude,
           date_time = GMT)
glimpse(whales)
```

```{r change date_time column class}
whales <- whales |> 
    mutate(date_time = dmy_hm(date_time))
glimpse(whales)
```

```{r set PTT as factor type}
# should only be 50 PTT numbers
unique(whales$PTT)

# change to factor type
whales <- whales |> 
    mutate(PTT = as.factor(PTT))
```

```{r make year, date, time columns}
whales <- whales |> 
    mutate(year = year(date_time),    # year for each GPS point
           date = date(date_time),    # date for each GPS point
           time = as_hms(date_time)) |>     # time for each GPS point
    relocate(year, .after = date_time) |> 
    relocate(date, .after = year) |> 
    relocate(time, .after = date)
```

```{r ARGOS location classes}
# change to factor type
whales <- whales |> 
    mutate(argos_loc_class = as.factor(argos_loc_class))

glimpse(whales$argos_loc_class)
```

ARGOS Location classes:
- A, B = no accuracy estimation
- 0 = estimated error > 1500m
- 1 = estimated error 500m < < 1500m
- 2 = estimated error 250 < < 500
- 3 = estimated error < 250 m

```{r}
glimpse(whales)
```


## Metadata

```{r}
metadata <- read_csv("C:/Users/jc984308/Downloads/AAS_4600_EastAus_humpback_satellite_tracking_data/AAS_4600_EastAus_humpback_satellite_tracking_data/4.Deploy_dets_HBW.csv")
#metadata <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/4.Deploy_dets_HBW.csv")
glimpse(metadata)

# rename columns
metadata <- metadata |> 
    rename(PTT = `Argos PTT`,
           dep_year = `Deploy year`,
           dep_date = `Deploy date`,
           end_date = `End date`,
           track_dur = `Tracking duration (days)`,
           dep_loc = `Deploy location`,
           dep_lat = `Deploy latitude`,
           dep_lon = `Deploy longitude`,
           stage = `Stage of annual cycle upon deployment`,
           sex = Sex,
           maturity = Maturity,
           init_activity = `Initial Activity`,
           num_locations = `Number of locations`,
           tag_prog = `Tag programming`,
           retain_SSM = `Retained for SSM`,
           SSM_est_dist = `SSM derived track distance estimate (km)`,
           movement = `Movement captured`)
glimpse(metadata)
```
```{r}
unique(metadata$dep_loc)  # change to factor
unique(metadata$stage)  # change to factor
unique(metadata$sex)  # change to factor
unique(metadata$maturity)   # change to factor
unique(metadata$init_activity)
unique(metadata$movement)
```

There are 50 PTT numbers in the data, each one representing a different humpback whale.

```{r change metadata date columns to POSIXct}
metadata <- metadata |> 
    mutate(dep_date = dmy(dep_date)) |> 
    mutate(end_date = dmy(end_date)) |>
    mutate(year = year(dep_year))

glimpse(metadata)
```

```{r change metadata PTT to factor type}
# change to factor type
metadata <- metadata |> 
    mutate(PTT = as.factor(PTT))

glimpse(metadata)
```

```{r change other variables to factor type}
metadata <- metadata |>
    mutate(dep_loc = as.factor(dep_loc)) |>
    mutate(stage = as.factor(stage)) |>
    mutate(sex = as.factor(sex)) |>
    mutate(maturity = as.factor(maturity))

glimpse(metadata)
```


```{r remove unnecessary metadata columns}
metadata_filt <- metadata |> 
    select(-c(14,15,16))

glimpse(metadata_filt)
```
```{r counts for factored variables}
metadata_filt
```


# Investigate the data

```{r merge dataframes}
whales_all <- whales |> 
    left_join(metadata_filt, by = join_by(PTT))

glimpse(whales_all)
```

**Is there a relationship between number of GPS points and where the whales went?**

```{r num GPS points per whale}
whales_all |> 
    group_by(PTT, dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |> 
    group_by(dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |> 
    group_by(PTT, init_activity) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |> 
    group_by(PTT, num_locations) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))
```

```{r}
ggplot(whales_all, aes(x = lon, y = lat, group = PTT, colour = PTT)) +
    geom_path() +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)")
```

```{r map all whales}
world <- ne_countries(scale = "medium", returnclass = "sf")

whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)") +
    theme(legend.position = "none")
```

```{r interactive map all whales}
# interactive map
p <- ggplot(whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
  geom_path()

ggplotly(p)
```

```{r separate map for each whale}
world <- ne_countries(scale = "medium", returnclass = "sf")
ptts <- unique(whales_all$PTT)

walk(ptts, function(p) {
  d <- whales_all %>% filter(PTT == p)
  
  # since dep_date/end_date should be constant per whale, grab the first value
  dep <- format(d$dep_date[1], "%d/%m/%Y")
  end <- format(d$end_date[1], "%d/%m/%Y")
  
  plt <- ggplot() +
      geom_sf(data = world, fill = "grey80", color = NA) +
      geom_path(data = d, aes(x = lon, y = lat, colour = "red")) +
      coord_sf(xlim = range(d$lon), ylim = range(d$lat)) +
      theme_minimal() +
      labs(title = paste("Whale PTT:", p), x = "Longitude", y = "Latitude",
           subtitle = paste("Tracked:", dep, " - ", end))
  
  print(plt)
})
```

```{r}
ptts <- unique(whales_all$PTT)

walk(ptts, function(p) {
  d <- whales_all %>% filter(PTT == p)
  
  dep <- format(d$dep_date[1], "%d/%m/%Y")
  end <- format(d$end_date[1], "%d/%m/%Y")
  
  plt <- ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = d, aes(x = lon, y = lat), color = "red", linewidth = 1.25) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(
      title = paste("Whale PTT:", p),
      subtitle = paste("Tracked:", dep, "-", end),
      x = "Longitude", y = "Latitude"
    )
  
  print(plt)
})
```

