---
title: "MB5370 R Project"
author: "Katie Dyck"
date: "2026-08-13"
format:
  html:
      df-scroll: true
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      code-overflow: scroll
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Housekeeping

Purge the global environment with the command `rm(list = ls())`.

```{r Clear session environment, echo=FALSE, include=FALSE}
#| output: false

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```

# Load packages

```{r load packages}
library(usethis)
library(tidyverse)
library(hms)
library(sf)
library(rnaturalearth)
library(plotly)
library(ggspatial)
library(maps)
```

```{r connect to GitHub on JCU computers, eval=FALSE}
# go to where you want the project stored (cd ~/Documents)
# in R terminal: git clone https://github.com/katiedyck/MB5370-Rproject.git
# move into folder: cd MB5370-Rproject
# after saving:
  # git add .
  # git commit -m "Comment"
  # git push

# if repository is already cloned:
  # cd ~/Documents/MB5370-RProject
  # git pull
  # git add .
  # git commit -m "Comment"
  # git push
```

# Data

## Tracking data

```{r import tracking data}
# whales <- read_csv("C:/Users/jc984308/Documents/MB5370/1.Argos_locs.csv")
whales <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/1.Argos_locs.csv")
glimpse(whales)
```
Rename columns for easier use.
```{r rename columns}
# rename columns
whales <- whales |> 
    rename(PTT = `Argos PTT`,
           argos_loc_class = `Argos location class`,
           lon = Longitude, lat = Latitude,
           date_time = GMT)
glimpse(whales)
```
Change `date_time` column to POSIXct class for easier manipulation.
```{r change date_time column class}
whales <- whales |> 
    mutate(date_time = dmy_hm(date_time))
glimpse(whales)
```
Set `PTT` column to factor type for easier manipulation.
```{r set PTT as factor type}
# should only be 50 PTT numbers
unique(whales$PTT)

# change to factor type
whales <- whales |> 
    mutate(PTT = as.factor(PTT))
```
Create a `year`, `date`, and `time` column.
```{r make year, date, time columns}
whales <- whales |> 
    mutate(year = year(date_time),    # year for each GPS point
           date = date(date_time),    # date for each GPS point
           time = as_hms(date_time)) |>     # time for each GPS point
    relocate(year, .after = date_time) |> 
    relocate(date, .after = year) |> 
    relocate(time, .after = date)
```

Change `argos_loc_class` column to factor type for easier manipulation.
```{r ARGOS location classes}
# change to factor type
whales <- whales |> 
    mutate(argos_loc_class = as.factor(argos_loc_class))

glimpse(whales$argos_loc_class)
```

ARGOS Location classes:
- A, B = no accuracy estimation
- 0 = estimated error > 1500m
- 1 = estimated error 500m < < 1500m
- 2 = estimated error 250 < < 500
- 3 = estimated error < 250 m

```{r glimpse data}
glimpse(whales)
```


## Metadata

```{r}
# metadata <- read_csv("C:/Users/jc984308/Documents/MB5370/4.Deploy_dets_HBW.csv")
metadata <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/4.Deploy_dets_HBW.csv")
glimpse(metadata)

# rename columns
metadata <- metadata |> 
    rename(PTT = `Argos PTT`,
           dep_year = `Deploy year`,
           dep_date = `Deploy date`,
           end_date = `End date`,
           track_dur = `Tracking duration (days)`,
           dep_loc = `Deploy location`,
           dep_lat = `Deploy latitude`,
           dep_lon = `Deploy longitude`,
           stage = `Stage of annual cycle upon deployment`,
           sex = Sex,
           maturity = Maturity,
           init_activity = `Initial Activity`,
           num_locations = `Number of locations`,
           tag_prog = `Tag programming`,
           retain_SSM = `Retained for SSM`,
           SSM_est_dist = `SSM derived track distance estimate (km)`,
           movement = `Movement captured`)
glimpse(metadata)
```
Look at the unique values for categorical variables in the metadata to see if they need to be changed to factor type.
```{r unique values for categorical variables}
unique(metadata$dep_loc)  # change to factor
unique(metadata$stage)  # change to factor
unique(metadata$sex)  # change to factor
unique(metadata$maturity)   # change to factor
unique(metadata$init_activity)  # change to factor
unique(metadata$movement)   # possibly change to factor - check for typos, if names can be shortened for clarity
unique(metadata$track_dur)
```

Edit the `movement` column to fix typos.
```{r stringr edit movement options}
metadata <- metadata |> 
    mutate(movement = str_replace_all(movement, "daat", "data"))

unique(metadata$movement) 
```

**There are 50 PTT numbers in the data, each one representing a different humpback whale.**

Change the `dep_date` and `end_date` columns to POSIXct class for easier manipulation.
```{r change metadata date columns to POSIXct}
metadata <- metadata |> 
    mutate(dep_date = dmy(dep_date)) |> 
    mutate(end_date = dmy(end_date))

glimpse(metadata)
```

Change the `PTT` column to factor type for easier manipulation.
```{r change metadata PTT to factor type}
# change to factor type
metadata <- metadata |> 
    mutate(PTT = as.factor(PTT))

glimpse(metadata)
```

Change other categorical variables to factor type for easier manipulation.
```{r change other variables to factor type}
metadata <- metadata |>
    mutate(dep_loc = as.factor(dep_loc)) |>
    mutate(dep_year = as.factor(dep_year)) |>
    mutate(stage = as.factor(stage)) |>
    mutate(sex = as.factor(sex)) |>
    mutate(maturity = as.factor(maturity)) |>
    mutate(init_activity = as.factor(init_activity)) |>
    mutate(movement = as.factor(movement))

glimpse(metadata)
```

Remove unnecessary metadata columns.
```{r remove unnecessary metadata columns}
metadata_filt <- metadata |> 
    select(-c(14,15,16))

glimpse(metadata_filt)
```


# Investigate the data

## Merge dataframes

Join the tracking data and metadata dataframes by the `PTT` column to create a new dataframe called `whales_all`. This will allow for easier analysis of the combined data.

```{r merge dataframes}
whales_all <- whales |> 
    left_join(metadata_filt, by = join_by(PTT))

glimpse(whales_all)

write.csv(whales_all, "all_whales_data.csv", row.names = FALSE)
```

Create a table of counts for each factor variable in the `whales_all` dataframe to see how many whales were tagged in each location, what their sex and maturity were, how many departed from each tagging location, and what their initial activity was. This will help to identify any patterns or trends in the data.

```{r counts for factor variables}
# argos location classes
argos_count <- whales_all |>
    group_by(argos_loc_class) |>
    summarise(count = n()) |>
    arrange(desc(count))
argos_count

# PTT
PTT_count <- whales_all |>
    group_by(PTT) |>
    summarise(count = n()) |>
    arrange(desc(count))
PTT_count

# dep_loc
dep_loc_count <- whales_all |> 
    group_by(dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))
dep_loc_count

# stage
stage_count <- whales_all |> 
    group_by(stage) |> 
    summarise(count = n()) |> 
    arrange(desc(count))
stage_count

# sex
sex_count <- whales_all |> 
    group_by(sex) |> 
    summarise(count = n()) |> 
    arrange(desc(count))
sex_count

# maturity
maturity_count <- whales_all |> 
    group_by(maturity) |> 
    summarise(count = n()) |> 
    arrange(desc(count))
maturity_count

# init_activity
initActivity_count <- whales_all |>
    group_by(init_activity) |>
    summarise(count = n()) |>
    arrange(desc(count))
initActivity_count

# movement
movement_count <- whales_all |>
    group_by(movement) |>
    summarise(count = n()) |>
    arrange(desc(count))
movement_count

# dep_year
depYear_count <- whales_all |>
    group_by(dep_year) |>
    summarise(count = n()) |>
    arrange(desc(count))
depYear_count
```

***Is there a relationship between number of GPS points and where the whales went?***

Determine the number of GPS points for each whale by deployment location, initial activity, and number of locations. Determine how many whales were tagged in each deployment location and what their initial activity was.
```{r num GPS points per whale}
PTT_depLoc <- whales_all |> 
    group_by(PTT, dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))
PTT_depLoc

PTT_initActivity <- whales_all |> 
    group_by(PTT, init_activity) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))
PTT_initActivity

PTT_numLocations <- whales_all |> 
    group_by(PTT, num_locations) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))
PTT_numLocations

depLoc_initActivity <- whales_all |>
    group_by(dep_loc, init_activity) |>
    summarise(n_points = n()) |>
    arrange(desc(n_points))
depLoc_initActivity
```

Create a dataframe for each PTT number to allow for easier analysis of each individual whale's tracking data.
```{r create dataframe for each PTT}
# Write a function that will create a dataframe for each PTT and save it as a dataframe/tibble (e.g. whale_96401)
# (created using Chattr)
create_whales_dataframes <- function() {
    ptt_values <- unique(whales_all$PTT)
    
    for (ptt_val in ptt_values) {
        filtered_data <- whales_all %>% 
            filter(PTT == ptt_val)
        df_name <- paste0("whale_", ptt_val)
        assign(df_name, filtered_data, envir = .GlobalEnv)
    }
}

create_whales_dataframes()
```

## Graph of Argos location classes

```{r argos graph}
ggplot(whales_all, aes(x = argos_loc_class, fill = argos_loc_class)) +
    geom_bar() +
    labs(x = "ARGOS Location Class", y = "Count",
         title = "ARGOS Location Classes for Humpback Whales, 2008-2010") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"), axis.title.x = element_text(vjust = -0.8),
          legend.position = "none") +
    geom_label(stat = "count", aes(label = after_stat(count)), vjust = -0.3, fill = "white", color = "black", label.size = 0.3) +
    ylim(0,6000)

ggsave("argos_location_classes.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

## Maps

Map the movement path of all 50 whales included in the data.

```{r map all whales}
world <- ne_countries(scale = "medium", returnclass = "sf")
whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    geom_point(data = tibble(lon = 149.9, lat = -37.066667), aes(x = lon, y = lat), color = "black", size = 2) +  # Eden
    geom_point(data = tibble(lon = 153.4333, lat = -29.116667), aes(x = lon, y = lat), color = "black", size = 2) +  # Evans Head
    geom_point(data = tibble(lon = 153.1, lat = -26.65), aes(x = lon, y = lat), color = "black", size = 2) +  # Sunshine Coast
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Movement Path of all Humpback Whales, 2008-2010") +
    theme(legend.position = "none", plot.title = element_text(face = "bold"))

ggsave("all_whales_movement_path.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

```{r map whales tagged in Eden, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

# create a dataframe for all whales tagged in Eden, Australia
eden_whales <- whales_all |>
  filter(dep_loc == "Eden, Australia") |>
  arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = eden_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    # add a point for Eden, Australia location
    geom_point(data = tibble(lon = 149.9, lat = -37.066667), aes(x = lon, y = lat), color = "black", size = 2) +
    coord_sf(xlim = range(eden_whales$lon), ylim = range(eden_whales$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Eden, Australia (n = 15)",
         subtitle = paste0("Tracking duration: ", min(eden_whales$dep_date), " - ", max(eden_whales$end_date))) +
    theme(plot.subtitle = element_text(face = "italic"), plot.title = element_text(face = "bold"))

ggsave("whales_tagged_in_Eden.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

```{r map whales tagged in Evans Head, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

# create a dataframe for all whales tagged in Evans Head, Australia
evansHead_whales <- whales_all |>
    filter(dep_loc == "Evans Head, Australia") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = evansHead_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    # add a dot for location of Evans Head, Australia
    geom_point(data = tibble(lon = 153.4333, lat = -29.116667), aes(x = lon, y = lat), color = "black", size = 3) +
    coord_sf(xlim = range(evansHead_whales$lon) + c(-6,2), ylim = range(evansHead_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Evans Head, Australia (n = 14)", 
         subtitle = paste0("Tracking duration: ", min(evansHead_whales$dep_date), " - ", max(evansHead_whales$end_date))) +
    theme(plot.subtitle = element_text(face = "italic"), plot.title = element_text(face = "bold"))

ggsave("whales_tagged_in_EvansHead.png", width = 6, height = 10, dpi = 300, path = "figures/")
```

```{r map whales tagged in Antarctica}
world <- ne_countries(scale = "medium", returnclass = "sf")

# create a dataframe for all whales tagged in Antarctica
antartcica_whales <- whales_all |>
    filter(dep_loc == "Antarctica") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = antartcica_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(antartcica_whales$lon) + c(-2,2), ylim = range(antartcica_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Antarctica (n = 14)",
         subtitle = paste0("Tracking duration: ", min(antartcica_whales$dep_date), " - ", max(antartcica_whales$end_date))) +
    theme(plot.subtitle = element_text(face = "italic"), plot.title = element_text(face = "bold"))

ggsave("whales_tagged_in_Antarctica.png", width = 10, height = 6, dpi = 300, path = "figures/")

# ggplot() +
#     geom_sf(data = world, fill = "grey70", color = NA) +
#     geom_path(data = antartcica_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
#     coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
#     theme_minimal() +
#     labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
#          title = "Whales tagged in Antarctica (n = 14)")
```

```{r map whales tagged in Sunshine Coast, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

# create a dataframe for all whales tagged in Sunshine Coast, Australia
sunshineCoast_whales <- whales_all |>
    filter(dep_loc == "Sunshine Coast, Australia") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = sunshineCoast_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    # add a point for Sunshine Coast, Australia location
    geom_point(data = tibble(lon = 153.1, lat = -26.65), aes(x = lon, y = lat), color = "black", size = 3) +
    coord_sf(xlim = range(sunshineCoast_whales$lon) + c(-6,8), ylim = range(sunshineCoast_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Sunshine Coast, Australia (n = 7)",
         subtitle = paste0("Tracking duration: ", min(sunshineCoast_whales$dep_date), " - ", max(sunshineCoast_whales$end_date))) +
    theme(plot.subtitle = element_text(face = "italic"), plot.title = element_text(face = "bold"))

ggsave("whales_tagged_in_SunshineCoast.png", width = 6, height = 10, dpi = 300, path = "figures/")
```

## Argos location classes

```{r map of all whales by argos loc class}
# map of whales by argos loc class
ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = argos_loc_class)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "ARGOS Location Class",
         title = "Movement of Humpback Whales by ARGOS Location Class") +
    theme(plot.title = element_text(face = "bold"))

ggsave("whales_by_argos_loc_class.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

Map all whales according to the stage of their migration movement (`stage` column).
```{r map whales by stage}
world <- ne_countries(scale = "medium", returnclass = "sf")
whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = stage)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Stage of Migration Movement",
         title = "Movement of Humpback Whales by Migration Stage") +
    theme(plot.title = element_text(face = "bold")) +
    scale_color_discrete(labels = c("Migrating north (n=1,251)", "Migrating south (n=10,845)", "On feeding grounds (n=5,010)"))

ggsave("whales_by_stage.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

Map all whales according to their sex (`sex` column).
```{r map whales by sex}
world <- ne_countries(scale = "medium", returnclass = "sf")
whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = sex)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Sex",
         title = "Movement of Humpback Whales by Sex") +
    theme(plot.title = element_text(face = "bold")) +
    scale_color_discrete(labels = c("Female (n = 8,278)", "Male (n = 6,824)", "Unknown (n = 2,004)"))

ggsave("whales_by_sex.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

Map all whales according to their deployment year (`dep_year` column).
```{r map whales by dep_year}
world <- ne_countries(scale = "medium", returnclass = "sf")
whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = dep_year)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Tag Deployment Year",
         title = "Movement of Humpback Whales by Tag Deployment Year") +
    theme(plot.title = element_text(face = "bold")) +
    scale_color_discrete(labels = c("2008 (n = 7,562)", "2009 (n = 1,251)", "2010 (n = 7,562)"))

ggsave("whales_by_dep_year.png", width = 10, height = 6, dpi = 300, path = "figures/")
```
Map whales according to their movement patterns (`movement` column).
```{r map whales by movement}
world <- ne_countries(scale = "medium", returnclass = "sf")

# create a new column called movement_short to shorten the movement patterns for easier visualization
whales_all |>
  mutate(
    movement_short = case_when(
      grepl("Antarctic feeding", movement) ~ "Feeding",
      grepl("Northern migration", movement) ~ "Northward",
      grepl("Southern migration", movement) ~ "Southward",
      grepl("New Zealand", movement) ~ "Via NZ",
      TRUE ~ "Other"
    )
  ) |>
    ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(aes(x = lon, y = lat, group = PTT, color = movement_short)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Movement",
         title = "Movement Patterns of Humpback Whales") +
    theme(plot.title = element_text(face = "bold")) +
    scale_color_discrete(labels = c("Feeding (n = 14,094)", "Northward (n = 1,249)", "Other (n = 395)", "Southward (n = 1,368)"))

ggsave("whales_by_movement.png", width = 10, height = 6, dpi = 300, path = "figures/")

movement_short_count <- whales_all |>
  mutate(
    movement_short = case_when(
      grepl("Antarctic feeding", movement) ~ "Feeding",
      grepl("Northern migration", movement) ~ "Northward",
      grepl("Southern migration", movement) ~ "Southward",
      grepl("New Zealand", movement) ~ "Via NZ",
      TRUE ~ "Other"
    )
  ) |> 
    group_by(movement_short) |>
    summarise(count = n()) |>
    arrange(desc(count))
movement_short_count
```

```{r interactive map all whales}
# interactive map
# p <- ggplot(whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
#   geom_path()
# 
# ggplotly(p)
```

```{r separate map for each whale}
# world <- ne_countries(scale = "medium", returnclass = "sf")
# ptts <- unique(whales_all$PTT)
# 
# walk(ptts, function(p) {
#   d <- whales_all %>% filter(PTT == p)
#   
#   # since dep_date/end_date should be constant per whale, grab the first value
#   dep <- format(d$dep_date[1], "%d/%m/%Y")
#   end <- format(d$end_date[1], "%d/%m/%Y")
#   
#   plt <- ggplot() +
#       geom_sf(data = world, fill = "grey80", color = NA) +
#       geom_path(data = d, aes(x = lon, y = lat, colour = "red")) +
#       coord_sf(xlim = range(d$lon), ylim = range(d$lat)) +
#       theme_minimal() +
#       labs(title = paste("Whale PTT:", p), x = "Longitude", y = "Latitude",
#            subtitle = paste("Tracked:", dep, " - ", end)) +
#       theme(legend.position = "none")
#   
#   print(plt)
# })
```

```{r separate map for each whale, with larger lines and full map}
# ptts <- unique(whales_all$PTT)
# 
# walk(ptts, function(p) {
#   d <- whales_all %>% filter(PTT == p)
#   
#   dep <- format(d$dep_date[1], "%d/%m/%Y")
#   end <- format(d$end_date[1], "%d/%m/%Y")
#   
#   plt <- ggplot() +
#     geom_sf(data = world, fill = "grey70", color = NA) +
#     geom_path(data = d, aes(x = lon, y = lat), color = "red", linewidth = 1.25) +
#     coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
#     theme_minimal() +
#     labs(
#       title = paste("Whale PTT:", p),
#       subtitle = paste("Tracked:", dep, "-", end),
#       x = "Longitude", y = "Latitude"
#     )
#   
#   print(plt)
# })
```

## Correlation between variables

```{r correlation bw dep_loc and init_activity}
# check correlation between dep_loc and init_activity
ggplot(whales_all, aes(x = dep_loc, fill = init_activity)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Initial Activity",
         title = "Association between Deployment Location and Initial Activity") +
    theme_minimal() +
    # rename fill legend labels
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    theme(axis.title.x = element_text(vjust = -0.8), plot.title = element_text(face = "bold"))
ggsave("depLoc_initActivity_correlation.png", width = 10, height = 6, dpi = 300, path = "figures/")

whales_all |>
    mutate(dep_loc = as.numeric(dep_loc),
           init_activity = as.numeric(init_activity)) |>
    summarise(correlation = cor(dep_loc, init_activity, use = "complete.obs"))
```

```{r correlation bw dep_loc and sex}
# check correlation between dep_loc and sex
ggplot(whales_all, aes(x = dep_loc, fill = sex)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Sex") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Female (n = 8278)", "Male (n = 6824)", "Unknown (n = 2004)"))

whales_all |>
    mutate(dep_loc = as.numeric(dep_loc),
           sex = as.numeric(sex)) |>
    summarise(correlation = cor(dep_loc, sex, use = "complete.obs"))
```

```{r correlation bw dep_loc and maturity}
# correlation between dep_loc and maturity
ggplot(whales_all, aes(x = dep_loc, fill = maturity)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Maturity") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Adult (n = 16,722)", "Subadult (n = 377)", "Calf (n = 7)"))

# 1. create contingency table
depLoc_matur_table <- table(whales_all$dep_loc, whales_all$maturity)

# 2. run chi-squared test
print(chisq.test(depLoc_matur_table))  # p-value < 0.05 = significant association between dep_loc and maturity
```

```{r correlation bw dep_loc and stage}
# correlation between dep_loc and stage
ggplot(whales_all, aes(x = dep_loc, fill = stage)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Stage of Migration Movement",
         title = "Association between Deployment Location and Migration Movement") +
    theme_minimal() +
    theme(axis.title.x = element_text(vjust = -0.8), plot.title = element_text(face = "bold")) +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Migrating South (n = 10,845)", "At Feeding Grounds (n = 5010)", "Migrating North (n = 1251)"))
ggsave("depLoc_stage_correlation.png", width = 10, height = 6, dpi = 300, path = "figures/")

# 1. create contingency table
depLoc_stage_table <- table(whales_all$dep_loc, whales_all$stage)

# 2. run chi-squared test
print(chisq.test(depLoc_stage_table))  # p-value < 0.05 = significant association between dep_loc and stage
```

```{r correlation bw sex and stage}
# correlation between sex and stage
ggplot(whales_all, aes(x = stage, fill = sex)) +
    geom_bar(position = "dodge") +
    labs(x = "Stage of Migration Movement", y = "Count", fill = "Sex",
         title = "Association between Sex and Stage of Migration Movement") +
    theme_minimal() +
    scale_x_discrete(labels = c("Migrating North\n(n = 1251)", "Migrating South\n(n = 10,845)", "On Feeding Grounds\n(n = 5010)")) +
    theme(axis.title.x = element_text(vjust = -0.8), plot.title = element_text(face = "bold"))
ggsave("sex_stage_correlation.png", width = 10, height = 6, dpi = 300, path = "figures/")

# 1. create contingency table
sex_stage_table <- table(whales_all$sex, whales_all$stage)

# 2. run chi-squared test
print(chisq.test(sex_stage_table))
```

```{r correlation bw dep_year and dep_loc}
# correlation between dep_year and dep_loc
ggplot(whales_all, aes(x = dep_year, fill = dep_loc)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Year", y = "Count", fill = "Deployment Location") +
    theme_minimal() +
    scale_fill_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU"))

ggplot(whales_all, aes(x = dep_loc, fill = dep_year)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Deployment Year",
         title = "Deployment Year by Deployment Location, 2008-2010") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    theme(plot.title = element_text(face = "bold"), axis.title.x = element_text(vjust = -0.8)) +
    # geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.8, fontface = "bold") +
    geom_label(stat = "count", aes(label = after_stat(count)), vjust = -0.3, fill = "white", color = "black", label.size = 0.3) +
    ylim(0,9000)
ggsave("depYear_depLoc_correlation.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

## Number of tracking points per whale

```{r}
max(PTT_count$count)  # max number of tracking points for a whale
min(PTT_count$count)  # min number of tracking points for a whale
```

```{r histrogram tracking ptts per whale}
ggplot(whales_all, aes(x = PTT)) +
    geom_bar(fill = "steelblue") +
    labs(x = "Whale (PTT)", y = "Number of Tracking Points",
         title = "Number of Tracking Points for each Humpback Whale", subtitle = "All Whales (n = 50)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic"), axis.title.x = element_text(vjust = -1))
ggsave("tracking_points_all_whales.png", width = 10, height = 6, dpi = 300, path = "figures/")

whales_all |>
    group_by(PTT) |>
    summarise(n_points = n()) |>
    filter(n_points > 250) |>
    ggplot(aes(x = PTT, y = n_points)) +
    geom_col(fill = "steelblue") +
    labs(x = "Whale (PTT)", y = "Number of Tracking Points",
         title = "Number of Tracking Points for each Humpback Whale", subtitle = "Individuals with > 250 tracking points (n = 19)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic"), axis.title.x = element_text(vjust = -1))
ggsave("tracking_points_250_plus.png", width = 10, height = 6, dpi = 300, path = "figures/")

whales_all |>
    group_by(PTT) |>
    summarise(n_points = n()) |>
    filter(n_points > 500) |>
    ggplot(aes(x = PTT, y = n_points)) +
    geom_col(fill = "steelblue") +
    labs(x = "Whale (PTT)", y = "Number of Tracking Points",
         title = "Number of Tracking Points for each Humpback Whales", subtitle = "Individuals with > 500 tracking points (n = 11)") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(face = "italic"), axis.title.x = element_text(vjust = -1))
ggsave("tracking_points_500_plus.png", width = 10, height = 6, dpi = 300, path = "figures/")

whales_all |>
    group_by(PTT) |>
    summarise(n_points = n()) |>
    filter(n_points > 1000) |>
    ggplot(aes(x = PTT, y = n_points)) +
    geom_col(fill = "steelblue") +
    labs(x = "Whale (PTT)", y = "Number of Tracking Points",
         title = "Number of Tracking Points for each Humpback Whale", subtitle = "Individuals with > 1000 tracking points (n = 5)") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(face = "italic"), axis.title.x = element_text(vjust = -1))
ggsave("tracking_points_1000_plus.png", width = 10, height = 6, dpi = 300, path = "figures/")

# most whales are tracked briefly!
```

## Tracking duration

```{r}
whales_all |>
    group_by(PTT, track_dur) |>
    summarise(count = n())
```

```{r tracking duration plot}
whale_summary <- whales_all %>%
  group_by(PTT) %>%
  summarise(track_dur = max(track_dur))

ggplot(whale_summary, aes(x = PTT, y = track_dur)) +
  geom_col(fill = "steelblue") +
  labs(x = "Whale (PTT)", y = "Tracking Duration (days)",
       title = "Tracking Duration for each Humpback Whale") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(face = "bold"),
        axis.title.x = element_text(vjust = -1))

ggsave("tracking_duration_per_whale.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

```{r}
# average track_dur
mean(whale_summary$track_dur)

# longest tracking duration
max(whale_summary$track_dur)

# shortest tracking duration
min(whale_summary$track_dur)
```

## Distribution of whale sex

```{r distribution of whale sex}
ggplot(whales_all, aes(sex, fill = sex)) +
    geom_bar() +
    theme_minimal() +
    labs(x = "Sex", y = "Count", title = "Distribution of Humpback Whale Sex") +
    theme(legend.position = "none", plot.title = element_text(face = "bold")) +
    # add a label with the count of each sex in each bar
    geom_label(stat = "count", aes(label = after_stat(count)), vjust = -0.3, fill = "white", color = "black", label.size = 0.3) +
    ylim(0, 9000)

ggsave("whales_by_sex_barplot.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

## Maturity of tracked whales

```{r maturity of whales}
ggplot(whales_all, aes(maturity, fill = maturity)) +
    geom_bar() +
    theme(legend.position = "none")

print(maturity_count)
```

## Movement patterns

```{r}
whales_all |>
    mutate(
        movement_short = case_when(
            grepl("Antarctic feeding", movement) ~ "Feeding",
            grepl("Northern migration", movement) ~ "Northward",
            grepl("Southern migration", movement) ~ "Southward",
            grepl("New Zealand", movement) ~ "Via NZ",
            TRUE ~ "Other"
            )
        ) |>    
    ggplot(aes(movement_short, fill = sex)) +
    geom_bar(position = "dodge") +
    theme_minimal() +
    labs(x = "Movement Pattern", y = "Count", fill = "Sex",
         title = "Movement Patterns of Humpback Whales, 2008-2010") +
    theme(axis.title.x = element_text(vjust = -1), plot.title = element_text(face = "bold"))

ggsave("whales_by_movement_pattern.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

## Deployment locations

```{r}
ggplot(whales_all, aes(dep_loc, fill = dep_loc)) +
    geom_bar() +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "Deployment Location", y = "Count", title = "Number of Tags Deployed at Each Location, 2008-2010") +
    theme(plot.title = element_text(face = "bold"), axis.title.x = element_text(vjust = -1)) +
    # add a label with the count of each deployment location in each bar
    geom_label(stat = "count", aes(label = after_stat(count)), vjust = -0.3, fill = "white", color = "black", label.size = 0.3) +
    ylim(0, 9000)

ggsave("whales_by_deployment_location.png", width = 10, height = 6, dpi = 300, path = "figures/")
```

