---
title: "MB5370 R Project"
author: "Katie Dyck"
date: "2026-08-13"
format:
  html:
      df-scroll: true
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      code-overflow: scroll
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Housekeeping

Purge the global environment with the command `rm(list = ls())`.

```{r Clear session environment, echo=FALSE, include=FALSE}
#| output: false

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```

# Load packages

```{r load packages}
library(usethis)
library(tidyverse)
library(hms)
library(sf)
library(rnaturalearth)
library(plotly)
```

```{r connect to GitHub on JCU computers, eval=FALSE}
# go to where you want the project stored (cd ~/Documents)
# in R terminal: git clone https://github.com/katiedyck/MB5370-Rproject.git
# move into folder: cd MB5370-Rproject
# after saving:
  # git add .
  # git commit -m "Comment"
  # git push

# if repository is already cloned:
  # cd ~/Documents/MB5370-RProject
  # git pull
  # git add .
  # git commit -m "Comment"
  # git push
```

# Data

## Tracking data

```{r import tracking data}
whales <- read_csv("C:/Users/jc984308/Documents/MB5370/1.Argos_locs.csv")
#whales <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/1.Argos_locs.csv")
glimpse(whales)
```

```{r rename columns}
# rename columns
whales <- whales |> 
    rename(PTT = `Argos PTT`,
           argos_loc_class = `Argos location class`,
           lon = Longitude, lat = Latitude,
           date_time = GMT)
glimpse(whales)
```

```{r change date_time column class}
whales <- whales |> 
    mutate(date_time = dmy_hm(date_time))
glimpse(whales)
```

```{r set PTT as factor type}
# should only be 50 PTT numbers
unique(whales$PTT)

# change to factor type
whales <- whales |> 
    mutate(PTT = as.factor(PTT))
```

```{r make year, date, time columns}
whales <- whales |> 
    mutate(year = year(date_time),    # year for each GPS point
           date = date(date_time),    # date for each GPS point
           time = as_hms(date_time)) |>     # time for each GPS point
    relocate(year, .after = date_time) |> 
    relocate(date, .after = year) |> 
    relocate(time, .after = date)
```

```{r ARGOS location classes}
# change to factor type
whales <- whales |> 
    mutate(argos_loc_class = as.factor(argos_loc_class))

glimpse(whales$argos_loc_class)
```

ARGOS Location classes:
- A, B = no accuracy estimation
- 0 = estimated error > 1500m
- 1 = estimated error 500m < < 1500m
- 2 = estimated error 250 < < 500
- 3 = estimated error < 250 m

```{r}
glimpse(whales)
```


## Metadata

```{r}
metadata <- read_csv("C:/Users/jc984308/Documents/MB5370/4.Deploy_dets_HBW.csv")
#metadata <- read_csv("~/Documents/JCU/2026 TR2/MB5370/R Project/AAS_4600_EastAus_humpback_satellite_tracking_data/4.Deploy_dets_HBW.csv")
glimpse(metadata)

# rename columns
metadata <- metadata |> 
    rename(PTT = `Argos PTT`,
           dep_year = `Deploy year`,
           dep_date = `Deploy date`,
           end_date = `End date`,
           track_dur = `Tracking duration (days)`,
           dep_loc = `Deploy location`,
           dep_lat = `Deploy latitude`,
           dep_lon = `Deploy longitude`,
           stage = `Stage of annual cycle upon deployment`,
           sex = Sex,
           maturity = Maturity,
           init_activity = `Initial Activity`,
           num_locations = `Number of locations`,
           tag_prog = `Tag programming`,
           retain_SSM = `Retained for SSM`,
           SSM_est_dist = `SSM derived track distance estimate (km)`,
           movement = `Movement captured`)
glimpse(metadata)
```
```{r unique values for categorical variables}
unique(metadata$dep_loc)  # change to factor
unique(metadata$stage)  # change to factor
unique(metadata$sex)  # change to factor
unique(metadata$maturity)   # change to factor
unique(metadata$init_activity)  # change to factor
unique(metadata$movement)
```

There are 50 PTT numbers in the data, each one representing a different humpback whale.

```{r change metadata date columns to POSIXct}
metadata <- metadata |> 
    mutate(dep_date = dmy(dep_date)) |> 
    mutate(end_date = dmy(end_date)) |>
    mutate(dep_year = year(dep_year))

glimpse(metadata)
```

```{r change metadata PTT to factor type}
# change to factor type
metadata <- metadata |> 
    mutate(PTT = as.factor(PTT))

glimpse(metadata)
```

```{r change other variables to factor type}
metadata <- metadata |>
    mutate(dep_loc = as.factor(dep_loc)) |>
    mutate(stage = as.factor(stage)) |>
    mutate(sex = as.factor(sex)) |>
    mutate(maturity = as.factor(maturity)) |>
    mutate(init_activity = as.factor(init_activity)) |>
    mutate(movement = as.factor(movement))

glimpse(metadata)
```

```{r remove unnecessary metadata columns}
metadata_filt <- metadata |> 
    select(-c(14,15,16))

glimpse(metadata_filt)
```


# Investigate the data

## Merge dataframes

```{r merge dataframes}
whales_all <- whales |> 
    left_join(metadata_filt, by = join_by(PTT))

glimpse(whales_all)
```

```{r counts for factor variables}
# argos location classes
whales_all |>
    group_by(argos_loc_class) |>
    summarise(count = n()) |>
    arrange(desc(count))

# PTT
whales_all |>
    group_by(PTT) |>
    summarise(count = n()) |>
    arrange(desc(count))

# dep_loc
whales_all |> 
    group_by(dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

# stage
whales_all |> 
    group_by(stage) |> 
    summarise(count = n()) |> 
    arrange(desc(count))

# sex
whales_all |> 
    group_by(sex) |> 
    summarise(count = n()) |> 
    arrange(desc(count))

# maturity
whales_all |> 
    group_by(maturity) |> 
    summarise(count = n()) |> 
    arrange(desc(count))

# init_activity
whales_all |>
    group_by(init_activity) |>
    summarise(count = n()) |>
    arrange(desc(count))

# movement
whales_all |>
    group_by(movement) |>
    summarise(count = n()) |>
    arrange(desc(count))
```

**Is there a relationship between number of GPS points and where the whales went?**

```{r num GPS points per whale}
whales_all |> 
    group_by(PTT, dep_loc) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |> 
    group_by(PTT, init_activity) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |> 
    group_by(PTT, num_locations) |> 
    summarise(n_points = n()) |> 
    arrange(desc(n_points))

whales_all |>
    group_by(dep_loc, init_activity) |>
    summarise(n_points = n()) |>
    arrange(desc(n_points))
```

## Maps

```{r map whales tagged in Eden, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

eden_whales <- whales_all |>
  filter(dep_loc == "Eden, Australia") |>
  arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = eden_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(eden_whales$lon), ylim = range(eden_whales$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Eden, Australia (n = 15)")
```

```{r map whales tagged in Evans Head, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

evansHead_whales <- whales_all |>
    filter(dep_loc == "Evans Head, Australia") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = evansHead_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(evansHead_whales$lon) + c(-6,2), ylim = range(evansHead_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Evans Head, Australia (n = 14)")
```

```{r map whales tagged in Antarctica}
world <- ne_countries(scale = "medium", returnclass = "sf")

antartcica_whales <- whales_all |>
    filter(dep_loc == "Antarctica") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = antartcica_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(antartcica_whales$lon) + c(-2,2), ylim = range(antartcica_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Antarctica (n = 14)")

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = antartcica_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Antarctica (n = 14)")
```

```{r map whales tagged in Sunshine Coast, Australia}
world <- ne_countries(scale = "medium", returnclass = "sf")

sunshineCoast_whales <- whales_all |>
    filter(dep_loc == "Sunshine Coast, Australia") |>
    arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = sunshineCoast_whales, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(sunshineCoast_whales$lon) + c(-6,8), ylim = range(sunshineCoast_whales$lat) + c(-2,2)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)",
         title = "Whales tagged in Sunshine Coast, Australia (n = 7)")
```

```{r map all whales}
world <- ne_countries(scale = "medium", returnclass = "sf")

whales_all <- whales_all %>% arrange(PTT, date_time)

ggplot() +
    geom_sf(data = world, fill = "grey70", color = NA) +
    geom_path(data = whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
    coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
    theme_minimal() +
    labs(x = "Longitude", y = "Latitude", color = "Whale (PTT)") +
    theme(legend.position = "none")
```

```{r interactive map all whales}
# interactive map
p <- ggplot(whales_all, aes(x = lon, y = lat, group = PTT, color = PTT)) +
  geom_path()

ggplotly(p)
```

```{r separate map for each whale}
# world <- ne_countries(scale = "medium", returnclass = "sf")
# ptts <- unique(whales_all$PTT)
# 
# walk(ptts, function(p) {
#   d <- whales_all %>% filter(PTT == p)
#   
#   # since dep_date/end_date should be constant per whale, grab the first value
#   dep <- format(d$dep_date[1], "%d/%m/%Y")
#   end <- format(d$end_date[1], "%d/%m/%Y")
#   
#   plt <- ggplot() +
#       geom_sf(data = world, fill = "grey80", color = NA) +
#       geom_path(data = d, aes(x = lon, y = lat, colour = "red")) +
#       coord_sf(xlim = range(d$lon), ylim = range(d$lat)) +
#       theme_minimal() +
#       labs(title = paste("Whale PTT:", p), x = "Longitude", y = "Latitude",
#            subtitle = paste("Tracked:", dep, " - ", end)) +
#       theme(legend.position = "none")
#   
#   print(plt)
# })
```

```{r separate map for each whale, with larger lines and full map}
# ptts <- unique(whales_all$PTT)
# 
# walk(ptts, function(p) {
#   d <- whales_all %>% filter(PTT == p)
#   
#   dep <- format(d$dep_date[1], "%d/%m/%Y")
#   end <- format(d$end_date[1], "%d/%m/%Y")
#   
#   plt <- ggplot() +
#     geom_sf(data = world, fill = "grey70", color = NA) +
#     geom_path(data = d, aes(x = lon, y = lat), color = "red", linewidth = 1.25) +
#     coord_sf(xlim = range(whales_all$lon), ylim = range(whales_all$lat)) +
#     theme_minimal() +
#     labs(
#       title = paste("Whale PTT:", p),
#       subtitle = paste("Tracked:", dep, "-", end),
#       x = "Longitude", y = "Latitude"
#     )
#   
#   print(plt)
# })
```

## Correlation between variables

```{r correlation bw dep_loc and init_activity}
# check correlation between dep_loc and init_activity
ggplot(whales_all, aes(x = dep_loc, fill = init_activity)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Initial Activity") +
    theme_minimal() +
    # rename fill legend labels
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU"))

whales_all |>
    mutate(dep_loc = as.numeric(dep_loc),
           init_activity = as.numeric(init_activity)) |>
    summarise(correlation = cor(dep_loc, init_activity, use = "complete.obs"))
```

```{r correlation bw dep_loc and sex}
# check correlation between dep_loc and sex
ggplot(whales_all, aes(x = dep_loc, fill = sex)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Sex") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Female (n = 8278)", "Male (n = 6824)", "Unknown (n = 2004)"))

whales_all |>
    mutate(dep_loc = as.numeric(dep_loc),
           sex = as.numeric(sex)) |>
    summarise(correlation = cor(dep_loc, sex, use = "complete.obs"))
```

```{r correlation bw dep_loc and maturity}
# correlation between dep_loc and maturity
ggplot(whales_all, aes(x = dep_loc, fill = maturity)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Maturity") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Adult (n = 16,722)", "Subadult (n = 377)", "Calf (n = 7)"))

# 1. create contingency table
depLoc_matur_table <- table(whales_all$dep_loc, whales_all$maturity)

# 2. run chi-squared test
print(chisq.test(depLoc_matur_table))  # p-value < 0.05 = significant association between dep_loc and maturity
```

```{r correlation bw dep_loc and stage}
# correlation between dep_loc and stage
ggplot(whales_all, aes(x = dep_loc, fill = stage)) +
    geom_bar(position = "dodge") +
    labs(x = "Deployment Location", y = "Count", fill = "Stage of Migration Movement") +
    theme_minimal() +
    scale_x_discrete(labels = c("Antarctica", "Eden, AU", "Evans Head, AU", "Sun. Coast, AU")) +
    scale_fill_discrete(labels = c("Migrating South (n = 10,845)", "At Feeding Grounds (n = 5010)", "Migrating North (n = 1251)"))

# 1. create contingency table
depLoc_stage_table <- table(whales_all$dep_loc, whales_all$stage)

# 2. run chi-squared test
print(chisq.test(depLoc_stage_table))  # p-value < 0.05 = significant association between dep_loc and stage
```

```{r correlation bw sex and stage}
# correlation between sex and stage
ggplot(whales_all, aes(x = stage, fill = sex)) +
    geom_bar(position = "dodge") +
    labs(x = "Stage of Migration Movement", y = "Count", fill = "Sex") +
    theme_minimal() +
    scale_x_discrete(labels = c("Migrating South\n(n = 10,845)", "At Feeding Grounds\n(n = 5010)", "Migrating North\n(n = 1251)"))
```

