---
title: "MB5370 R Project"
author: "Katie Dyck"
date: "2026-08-13"
format:
  html:
      df-scroll: true
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      code-overflow: scroll
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Housekeeping

Purge the global environment with the command `rm(list = ls())`.

```{r Clear session environment, echo=FALSE, include=FALSE}
#| output: false

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```

# Load packages

Packages necessary for my analysis are `tidyverse`, `RColorBrewer`, `sf`, `rnaturalearth`, and `rnaturalearthdata`. The `sf` package is used for working with spatial data. The `rnaturalearth` and `rnaturalearthdata` packages are used for creating basemaps.

```{r load packages, echo=FALSE, include=FALSE}
library(tidyverse)
library(sf)
library(RColorBrewer)   # high-contrast colour palette for mapping
library(rnaturalearth)   # basemap
library(rnaturalearthdata)
```

# Import data

Datasets were downloaded from the OBIS-SEAMAP database: https://seamap.env.duke.edu/dataset/2385.

There are 2 datasets available: one contains information on the paths each turtle being track travelling throughout the study period (`turtles_lines`), and the other contains information on the locations of each turtle at specific points in time (`turtles_points`).

```{r import data}
# Import the turtle lines data
turtles_lines <- read_csv("../../R Project/obis_seamap_dataset_2385_csv_v1.0.0_99185/obis_seamap_dataset_2385_lines.csv")
glimpse(turtles_lines)

# Import the turtle points data
turtles_points <- read_csv("../../R Project/obis_seamap_dataset_2385_csv_v1.0.0_99185/obis_seamap_dataset_2385_points.csv")
glimpse(turtles_points)
```

# Clean the datasets

## `turtle_lines` dataset

I will first look at some columns I am unsure of to determine if they will be relevant to my analysis.

```{r investigate lines columns, include=FALSE}
unique(turtles_lines$dataset_id)    # will remove this column because it is the same for all rows
head(unique(turtles_lines$row_id))    # will remove this column - just a unique identifier for each row
unique(turtles_lines$series_id) # this is the unique identifier for each of the 9 turtles, so I will keep this column
unique(turtles_lines$eff_status)    # will remove this column because it is the same for all rows, not sure what it means
unique(turtles_lines$ds_type)   # will remove this column because it is the same for all rows
unique(turtles_lines$id_obs)    # just the observer ID number, will remove because irrelevant to my analysis
```

Now I will remove unnecessary columns from the `turtles_lines` dataset.

```{r filter lines dataset}
turtles_lines_filtered <- turtles_lines %>% 
    select(-c(dataset_id, row_id, eff_status, ds_type, id_obs))
glimpse(turtles_lines_filtered)
```

I will rename the `series_id` column to `turtle_id` to make it more clear that this is the unique identifier for each turtle.

```{r rename lines series_id column}
turtles_lines_filtered <- turtles_lines_filtered %>%
    rename(turtle_id = series_id)
```

## `turtles_points` dataset

Will look at columns and remove those that are unnecessary as I did for the `turtles_lines` dataset.

```{r investigate points columns, include=FALSE}
unique(turtles_points$species_name)   # will remove this column because it is the same for all rows
unique(turtles_points$common_name)  # will remove this column because it is the same for all rows
unique(turtles_points$scientific_name)  # will keep this column
unique(turtles_points$itis_tsn)  # will remove this column because it is the same for all rows
unique(turtles_points$group_size)  # will remove this column because it is the same for all rows
unique(turtles_points$timezone)  # will remove this column because it is NA for all rows
unique(turtles_points$ds_type)  # will remove this column because it is NA for all rows
unique(turtles_points$platform)  # will remove this column because it is NA for all rows
unique(turtles_points$notes)  # will remove this column because it is NA for all rows
```

Need to change the `date_time` column from character to POSIXct type.

```{r change date_time to factored}
head(turtles_points$date_time)

turtles_points <- turtles_points %>%
    mutate(date_time = ymd_hm(date_time))

head(turtles_points$date_time)
str(turtles_points$date_time)
```

```{r filter points dataset}
# Remove unnecessary columns from `turtles_points`
turtles_points_filtered <- turtles_points %>% 
    select(-c(dataset_id, row_id, species_name, common_name, itis_tsn, group_size, timezone, ds_type, platform, provider, notes))
glimpse(turtles_points_filtered)
```

```{r rename points series_id column}
turtles_points_filtered <- turtles_points_filtered %>%
    rename(turtle_id = series_id) %>%
    relocate(turtle_id, .before = latitude) # move the turtle_id column to the front of the dataset

glimpse(turtles_points_filtered)
```

```{r check points date_time structure}
# check that the structure of date_time column in turtles_points_filtered is consistent
turtles_points_filtered %>%
    group_by(turtle_id) %>%
    summarise(
        min_date = min(date_time, na.rm = TRUE),
        max_date = max(date_time, na.rm = TRUE)
    )
```

* The `lprecision` and `tprecision` columns refer to the accuracy of the location and time data, respectively. They were added by OBIS-SEAMAP when the data was uploaded. I will keep these columns for now, but may remove them later if they are not relevant to my analysis.
* The `oceano` column was added by OBIS-SEAMAP; it was not included in the original dataset. It contains satellite-derived environmental data for each turtle location, including sea surface temperature, chlorophyll-a concentration, and sea surface height. I will keep this column for now, but may remove it later if it is not relevant to my analysis.

# What is the difference between the datasets?

The `turtles_lines` dataset contains information on the paths each turtle being tracked travelled throughout the study period, while the `turtles_points` dataset contains information on the locations of each turtle at specific points in time.

The `turtles_lines` dataset is a summary of the `turtles_points` dataset, with each row representing a unique turtle and its path over time.

The `turtles_points` dataset contains more detailed information on each turtle's location at specific points in time.

* Turtle 101010 was tracked from 05-03-2012 to 30-05-2012.
* Turtle 107822 was tracked from 10-01-2015 to 20-05-2015.
* Turtle 107833 was tracked from 15-12-2013 to 05-05-2014.
* Turtle 107844 was tracked from 09-12-2013 to 03-07-2014.
* Turtle 107893 was tracked from 14-12-2013 to 19-03-2014. 
* Turtle 107894 was tracked from 14-12-2012 to 14-01-2013.
* Turtle 107896 was tracked from 09-03-2015 to 30-03-2015. 
* Turtle 107898 was tracked from 23-01-2014 to 23-03-2014.
* Turtle 107899 was tracked from 23-01-2014 to 02-05-2014.

**INCLUDE A TABLE OF # OF TRACKING POINTS, START AND END DATES FOR EACH TURTLE IN PRESENTATION!!!**

# Mapping the turtle locations

## Prep the data

```{r prep data for mapping}
turtles <- turtles_points_filtered %>%
    mutate(
        turtle_id = factor(turtle_id)   # drop any unused levels
    ) %>%
    arrange(turtle_id, date_time) %>%
    filter(!is.na(longitude), !is.na(latitude), !is.na(date_time))

# should be 0 — that means all date/time formats have been accounted for 
sum(is.na(turtles$date_time))
 
# check: how many turtles, how many fixes each
turtles %>% count(turtle_id, sort = TRUE)
```

## Convert points to sf, then build one linestring per turtle

First, we need to convert points to a `sf` object, then build one linestring per turtle from the start and end points of each segment.

```{r convert points and build linestring}
# Convert to a sf object based on longitute and latitude columns
turtle_pts <- st_as_sf(turtles, coords = c("longitude", "latitude"), crs = 4326)
 
turtle_tracks <- turtle_pts %>%
    group_by(turtle_id) %>%
    summarise(do_union = FALSE) %>%   # do_union=FALSE preserves point order
    st_cast("LINESTRING")
```

```{r set levels and create colour palette}
# Set the levels of the turtle_id factor to ensure consistent color mapping across plots
turtle_ids <- levels(turtle_pts$turtle_id)

# Set a color palette for the turtles using RColorBrewer to ensure consistent colours across plots
turtle_palette <- setNames(brewer.pal(length(turtle_ids), "Set1"), turtle_ids)
```

## Create basemap

Create a basemap of Central America to plot the turtle locations and tracks on:

```{r create basemap}
# Basemap of Central America
land <- ne_countries(scale = "medium", returnclass = "sf")  # get polygons of countries; 'sf' = simple feature

bbox <- st_bbox(turtle_pts)
pad <- 1    # degree of padding around the tracks
```

## Build maps

```{r all turtles on one map}
p_all <- ggplot() +
    # Add basemap
    geom_sf(data = land, fill = "grey90", color = "grey60") +
    # Add turtle tracks
    geom_sf(data = turtle_tracks, aes(color = turtle_id), linewidth = 0.6) +
    # Add turtle tracking points
    geom_sf(data = turtle_pts, aes(color = turtle_id), size = 0.5, alpha = 0.5) +
    # Choose map coordinates
    coord_sf(
        xlim = c(bbox["xmin"] - pad, bbox["xmax"] + pad),
        ylim = c(bbox["ymin"] - pad, bbox["ymax"] + pad)
    ) +
    # Set colours for each turtle ID
    scale_color_manual(values = turtle_palette, name = "Turtle ID") +
    labs(
        title = "Green sea turtle tracks — Costa Rica (2012-2015)",
        caption = "OBIS-SEAMAP dataset 2385",
        x = NULL, y = NULL
    ) +
    theme_minimal()
 
p_all

# ggsave("turtle_tracks_all.png", p_all, width = 8, height = 10, dpi = 300)
```

For turtle ID 107822 and 107833, some of the tracking points are on land, far away from the shore, which is impossible. This is likely due to the fact that the tracking points are not accurate, and the turtle was likely in the water at the time. I will find these points in the dataset and exclude them from my analysis.

```{r}
# Find tracking points that are on land for turtles 107822 and 107833
land_points_107822 <- turtle_pts %>%
    filter(turtle_id == "107822") %>%
    st_join(land, join = st_intersects) %>%
    filter(!is.na(name))  # points that intersect with land polygons

land_points_107833 <- turtle_pts %>%
    filter(turtle_id == "107833") %>%
    st_join(land, join = st_intersects) %>%
    filter(!is.na(name))  # points that intersect with land polygons3

print(land_points_107822)
print(land_points_107833)
```




```{r facet_wrap map}
p_facet <- ggplot() +
    geom_sf(data = land, fill = "grey90", color = "grey60") +
    geom_sf(data = turtle_tracks, aes(color = turtle_id), linewidth = 0.6, show.legend = FALSE) +
    geom_sf(data = turtle_pts, size = 0.4, alpha = 0.5, show.legend = FALSE) +
    coord_sf(
        xlim = c(bbox["xmin"] - pad, bbox["xmax"] + pad),
        ylim = c(bbox["ymin"] - pad, bbox["ymax"] + pad)
    ) +
    facet_wrap(~turtle_id) +
    scale_color_manual(values = turtle_palette) +
    labs(title = "Individual turtle tracks", x = NULL, y = NULL) +
    theme_minimal()
 
p_facet
 
# ggsave("turtle_tracks_facets.png", p_facet, width = 10, height = 9, dpi = 300)
```

```{r separate map for each turtle ID}
turtle_plots <- map(turtle_ids, function(id) {
 
    pts_id <- turtle_pts %>%
        filter(turtle_id == id)
 
    track_id <- pts_id %>%
        summarise(do_union = FALSE) %>%
        st_cast("LINESTRING")
 
    bbox_id <- st_bbox(pts_id)
 
    date_range <- range(pts_id$date_time, na.rm = TRUE)
    date_label <- paste(
        format(date_range[1], "%d %b %Y"), "–", format(date_range[2], "%d %b %Y")
    )
 
    this_color <- turtle_palette[[id]]
 
    ggplot() +
        geom_sf(data = land, fill = "grey90", color = "grey60") +
        geom_sf(data = track_id, color = this_color, linewidth = 0.6) +
        geom_sf(data = pts_id, color = this_color, size = 0.8, alpha = 0.6) +
        coord_sf(
            xlim = c(bbox_id["xmin"] - pad, bbox_id["xmax"] + pad),
            ylim = c(bbox_id["ymin"] - pad, bbox_id["ymax"] + pad)
        ) +
        labs(
            title = paste("Turtle", id),
            subtitle = date_label,
            x = NULL, y = NULL
        ) +
        theme_minimal()
})
 
names(turtle_plots) <- turtle_ids
 
# display each map (in RStudio's plot pane / rendered document)
walk(turtle_plots, print)
 
# save each map to its own PNG, e.g. turtle_tracks_101010.png, etc.
iwalk(turtle_plots, ~ ggsave(
    filename = paste0("turtle_tracks_", .y, ".png"),
    plot = .x, width = 8, height = 8, dpi = 300
))
```

```{r interactive map view}
# library(mapview)
# mapview(turtle_tracks, zcol = "turtle_id") + mapview(turtle_pts, zcol = "turtle_id", cex = 2)
```

